Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-05-27 results.tex
updates

\section{Data illustrations\label{sec:results}}

\subsection{Simulations\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_p^{d-1}$. The generation procedure is detailed 
    in Algorithm~\ref{algo:simulated}.

\begin{algorithm}[ht]  % algorithm using algorithm package
        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
        \begin{algorithmic}
        \For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
        \For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
            \For{$j$ in $1,\ldots,n_{\text{mix}}$}
                \State Generate $\mu_{j} \sim \mathcal{N}\left(\bm{0},\bm{I}\right)$
                \State Generate $\Sigma_{j}\sim\mathcal{IW}\left(70,70 \bm{I}\right)$
            \EndFor
            \State Generate $\pi\sim\text{Dirichlet}(\bm{10})$
            \For{$i$ in $1,\ldots,1000$}
                \State Generate $\delta_i \sim \text{Categorical}(\pi)$
                \State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
            \EndFor
            \For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
            \EndFor
        \EndFor
        \EndFor
        \end{algorithmic}
    \end{algorithm}

where $\mu_j$, $\Sigma_j$ the parameters of the log-normal mixture component 
    distribution; $\pi$ the probability vector assigning weight to each mixture 
    component, and $\delta_i$ the mixture component identifier for each simulated 
    observation.

Figure~\ref{fig:simpples}, shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  We considered a total of four variations of the Dirichlet-process mixture
    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
    developed in~\cite{COOLEY2010}.  Note that the pairwise betas model respects
    the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    The point I am trying to make is that because the moment-restriction assumption is 
    discarded, a moment-restricted model is inadequate, as is borne out in the
    simulation study.}  We observe that the projected gamma models dominate, and within
    that class of model, the choice of centering distribution is irrelevant.  For the
    projected \emph{restricted} gamma models, the choice of log-normal centering
    distribution induces

shows two sets of panels that illustrate the comparison
between four different models used to fit the simulated data. We produced 16 
different data sets, one for each of four different numbers of mixture components 
and four different dimensions. We considered three different models: DP mixtures 
of projected gamma; Projected restricted gamma; and projected restricted gamma 
with a multivariate log-normal prior. We calculated $S^\infty_{\text{PPL}}$ and 
$S_{\text{ES}}$ using predictive samples from each of fitted models. To provide 
a comparative baseline, we calculated the scores using samples generated 
directly using the same model and parameters that produced the simulated 
observations. The results are denoted as \emph{Gen} in the figure. 
We observe that the projected
  restricted Gamma model dominates in most situations, but the difference between projected restricted
  Gamma and the other gamma-based models tends to shrink as the number of dimensions increases.
  Alternatively, that difference appears to grow as the number of mixture components increases. 
  The results from both scoring criteria are comparable.

@@

\begin{table}[htb]
  \centering
  \caption{Model comparison metrics: Posterior~Predictive~Loss~($S_{\text{PPL}}$) and 
    Energy~Score~($S_{\text{ES}}$) criteria from fitted models against the IVT data.  
    Lower is better. \label{tab:dev}}
  \include{./tables/table_dev}
\end{table}

Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy scores for the different models considered. We observe that, as in the case of the simulation study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
  gamma models.  Penalization for the unrestricted gamma model is 
  stronger for the real data than for the simulated data. We also observe that
  the restricted gamma model with the log-normal prior performs significantly better 
  than the restricted gamma model with the gamma prior on the higher dimensional data. This is the opposite of the result obtained in the simulation example.

>>>>

\section{Data illustrations\label{sec:results}} 
\makenote{this title seems weird?}

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
    procedure is detailed in Algorithm~\ref{algo:simulated}.  For each
    model, the data is then projected onto the appropriate $\mathbb{S}_p^{d-1}$
    for model fitting--dirichlet and pairwise betas being fit on 
    $\mathbb{S}_1^{d-1}$, and the projected gamma models being fit on 
    $\mathbb{S}_{10}^{d-1}$.

\begin{algorithm}[h]  % algorithm using algorithm package
        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
        $\mu_j$, $\Sigma_j$ are the parameters of the mixture component distribution; 
        $\pi$ is the probability vector assigning weight mixture components; $\delta_i$ 
        is the mixture component identifier for each simulated observation.}
        \begin{algorithmic}
        \For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
        \For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
            \For{$j$ in $1,\ldots,n_{\text{mix}}$}
                \State Generate $\mu_{j} \sim \mathcal{N}_{32}\left(\bm{0},\bm{I}\right)$
                \State Generate $\Sigma_{j}\sim\mathcal{IW}_{32}\left(70,70 \bm{I}\right)$
            \EndFor
            \State Generate $\pi\sim\text{Dirichlet}(\bm{10})$
            \For{$i$ in $1,\ldots,1000$}
                \State Generate $\delta_i \sim \text{Categorical}(\pi)$
                \State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
            \EndFor
            \For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
            \EndFor
        \EndFor
        \EndFor
        \end{algorithmic}
    \end{algorithm}

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  We considered a total of four variations of the Dirichlet-process mixture
    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
    developed in~\cite{COOLEY2010}.  Note that the pairwise betas model respects
    the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    The point I am trying to make is that because the moment-restriction assumption is 
    discarded, a moment-restricted model is inadequate, as is borne out in the
    simulation study.}  We observe that the projected gamma models dominate, and within
    that class of model, the choice of centering distribution is irrelevant.  For the
    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
    centering distribution negatively affects model fidelity.

@@

\begin{table}[htb]
  \centering
  \caption{Energy score criterion from fitted models against the IVT data. 
    Lower is better.\label{tab:dev}}
  \include{./tables/table_dev_updated}
\end{table}

Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
    scores for the different models considered. We observe that, as in the case of the simulation 
    study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
    Interestingly, this preference for the restricted gamma form is the opposite of the result 
    obtained in the simulation example.
